Draft Article

### 2019-10-30 article.tex
theory basics and flow update

\section{Theory}
One of the most prevalent clich\'es in spy literature and journalism is the so-called \enquote{rules of the game.} Media-friendly lists of rules like \enquote{You are never alone} and \enquote{Don't trust anyone} make the spy world seem insular, mysterious, and impenetrable.\footcite{myre_moscow_2019} They hint at a larger set of processes whose motivations the average citizen, not given access to this privileged information, could not possibly begin to understand. In reality, while the \emph{operations} of a modern intelligence service are incredibly opaque, the geopolitical conditions in which it operates are not. Every aspect of a nation-state's interactions with another ought to be subject to the same critical analysis. Scholars can use rationalist principles to analyze everything from trade policy to the space race; espionage, which by definition involves deeply invasive interactions with another state, ought to be no different.

Espionage certainly seems like something that states should be doing more to discourage. Then results of intelligence operations can have devastating, even existential effects on the targeted nation's security. One of the most traditional form of espionage is stealing military secrets from a foreign power, and in that arena China has been relentless. A US Navy report from March 2019 declared Chinese intelligence operations so extensive that they had substantively altered the balance of power between the two states.\footcite{lubold_navy_2019} ``Long-term, US future military advantage is being diminished by years of IP exfiltration from the DoD, DoN, and DIB,'' the report read, ``all with little to no adverse consequences to the thieves.''\footcite[p.~6]{bayer_cybersecurity_2019} The Navy's forecasts were dire: ``If the current trend continues unimpeded, the US will soon lose its status as the dominant global economic power.''\footcite[p.~5]{bayer_cybersecurity_2019} The Navy is not a disinterested party, of course, and there are reasons to take its conclusions with a grain of salt; modern military technology is remarkably complex and China cannot simply steal its way to military dominance.\footnote{For a good discussion of these limitations, see \cite{gilli_why_2019}} Nonetheless, the possibilities are alarming and the consequences for remain minimal.

>>>>

\section{Theory}
\subsection{Establishing the puzzle}
One of the most prevalent clich\'es in spy literature and journalism is the so-called \enquote{rules of the game.} Media-friendly lists of rules like \enquote{You are never alone} and \enquote{Don't trust anyone} make the spy world seem insular, mysterious, and impenetrable.\footcite{myre_moscow_2019} They hint at a larger set of processes whose motivations the average citizen, not given access to this privileged information, could not possibly begin to understand. In reality, while the \emph{operations} of a modern intelligence service are incredibly opaque, the geopolitical conditions in which one operates are not. Every aspect of a nation-state's interactions with another ought to be subject to the same critical analysis. Scholars can use rationalist principles to analyze everything from trade policy to the space race; espionage, which by definition involves deeply invasive interactions with another state, ought to be no different.

% Certainly there are factors that the go into diplomatic decisions that the general public will likely never know. There are few reasons to believe, however, that this is more true for intelligence-related diplomatic matters than for generic ones. Every diplomatic decision is made with the backing of classified intelligence.

Under traditional methods of analysis, espionage certainly seems like something that states should be doing more to discourage. Then results of intelligence operations can have devastating, even existential effects on the targeted nation's security. One of the most traditional form of espionage is stealing military secrets from a foreign power, and in that arena China has been relentless. A US Navy report from March 2019 declared Chinese intelligence operations so extensive that they had substantively altered the balance of power between the two states.\footcite{lubold_navy_2019} ``Long-term, US future military advantage is being diminished by years of IP exfiltration from the DoD, DoN, and DIB,'' the report read, ``all with little to no adverse consequences to the thieves.''\footcite[p.~6]{bayer_cybersecurity_2019} The Navy's forecasts were dire: ``If the current trend continues unimpeded, the US will soon lose its status as the dominant global economic power.''\footcite[p.~5]{bayer_cybersecurity_2019} The Navy is not a disinterested party, of course, and there are reasons to take its conclusions with a grain of salt; modern military technology is remarkably complex and China cannot simply steal its way to military dominance.\footnote{For a good discussion of these limitations, see \cite{gilli_why_2019}} Nonetheless, the possibilities are alarming and the consequences for remain minimal.

Not only does espionage have potentially disastrous consequences but the system is set up so that the structure of civilian intelligence services is never fundamentally threatened. The low percentage of responses for cyber-espionage have their basis in a similarly low response rate for traditional espionage. The traditional punishment for espionage is to declare certain embassy officials \emph{persona non grata}, which often results in ``tit-for-tat'' diplomatic expulsions. The name is taken from a highly-successful strategy for the prisoner's dilemma wherein the player simply mimics the choice that that their opponent made in the last round. By employing the tit-for-tat strategy, the player never gets taken advantage of for more than one turn and can always return to cooperation if their opponent unilaterally signals a willingness to do the same.

Diplomatic expulsions play out in exactly the same fashion: when one country expels the diplomats of another, the other country will often respond in kind. Each side will continue to expel each other's diplomats until one side signals that it would like for the reciprocal PNG-ing of diplomats to stop. Then, both sides return to a cooperative state, each preserving some of their intelligence operations. Neither side can really gain a decisive advantage, so the back-and-forth punishment never ends in a mutual defection. It might seem obvious to note, but \emph{both} states must have decided that preserving this system is valuable for them, otherwise just one of them refusing to cooperate would result in the mutual devastation of human intelligence operations.

\subsection{Espionage norms as a discrete series of choices}
Recall: the key feature of espionage norms is that the scope of the punishment is structurally incapable of discouraging a future, identical operation. Expelling suspected intelligence officers from an embassy temporarily makes it more difficult to conduct espionage, but it doesn't make a state less likely to try; the worst that can happen is a net-neutral result---going back to \enquote{square one.} By contrast, were it more common to respond to espionage by imposing financial sanctions or even inflicting a military cost, a state would have to consider the possibility that attempting espionage could adversely impact other aspects of their security, leaving it worse off than when it started. Such a response would be a violation of the well-established norms of espionage and it is essentially unheard-of.

Therefore, I propose that an iterated model can describe how states cooperate to preserve the liminal legal status of espionage. My model changes the parameters from above so that the decision to cooperate or defect is based not in the amount of punishment received (essentially measured by the number of diplomats expelled), but rather the type. Each time a state is forced to respond to a successful intelligence operation by a rival---such as the reveal of a high-level spy or a major data breach---it can either limit its response to that which is traditionally associated with espionage, or it can impose unprecedented costs that might discourage such an operation in the future. Table \ref{espionage-matrix} describes a single instance of this choice. States signal their interest in cooperating by limiting their retaliation in response to espionage, no matter how damaging the operation is to their national security. Each instance is a chance to play the game with fresh information; each espionage case is an opportunity to cooperate, reaffirming the norm that states always respond to espionage in a specific way, or defect, adding an additional cost to the attempt.

% https://tex.stackexchange.com/questions/249480/creating-a-payoff-matrix-using-latex-tabular-environment
\begin{table}[ht]
\centering
\setlength{\extrarowheight}{2pt}
\small
\begin{tabular}{cc|c|c|}
  & \multicolumn{1}{c}{} & \multicolumn{2}{c}{USSR}\\
  & \multicolumn{1}{c}{} & \multicolumn{1}{c}{Tolerate Espionage}  & \multicolumn{1}{c}{Punish Espionage} \\\cline{3-4}
  \multirow{3}*{USA}  & Tolerate Espionage & \makecell{~\\Both sides gain intelligence \\~} & \makecell{~\\ Only the USSR continues spying \\ ~} \\\cline{3-4}
  & Punish Espionage & \makecell{~\\ Only the US continues spying \\~} & \makecell{~\\ No intelligence either direction \\~} \\\cline{3-4}
\end{tabular}
\caption{Espionage consequences decision matrix}
\label{espionage-matrix}
\end{table}

The purpose of analyzing espionage norms this way is to remove the mystique.

Even though the US and the USSR have the opportunity to defect by imposing harsher punishments in response to espionage, the gains would be short-lived, because the other state is guaranteed to immediately do the same---tit-for-tat. Espionage consequences are an iterated game that both parties are playing the exactly the same way, so the only two possible outcomes long-term are mutual defection or mutual cooperation.

Given those two options, it makes intuitive sense that in the case of Cold War espionage both the US and the USSR chose to cooperate. Both sides gained a great deal from their intelligence operations; Soviet intelligence had an easier time operating in Washington than the US did in the USSR, but American intelligence was still formidable and Soviet defections were frequent.\footcite[p.~380]{macrakis_technophilic_2010} When the relative capabilities are not too lopsided, there is a pleasing reciprocity to espionage that makes cooperation easier.
